---
title: "Homework 4"
author: "Taylor Booth"
date: "March 16, 2021"
output: html_document
---

1. Fit a poisson regression model that assumes expected count is an interactive function of variables x1 and x2

```{r, echo = T}
dat <- read.csv("Homework 4 Data.csv")
head(dat)

fit <- glm(y ~ x1*x2, family = poisson, data = dat)
summary(fit)
```


2. Interpret the effect of variable x1 on the expected count when x2 is fixed at level 'b' verify your interpretation in R.


Answer: When x2 is fixed at b, then x1 has a log difference of 0.379

```{r, echo = T}
betas <- coef(fit)
betas

cnt_x1 <- betas[1] + betas[3]
cnt2_x1 <- betas[1] + betas[2] + betas[3] + betas[4]

log(exp(cnt2_x1) / exp(cnt_x1))

betas[2] + betas[4]
```


3. Interpret the effect of variable x2 on the expected count when x1 is fixed at 1. Verify your interpretation in R.


Answer: x2 has a log difference of -0.578 when x1 is fixed t 1

```{r, echo = T}
cnt_x2 <- betas[1] + betas[2] * 1 
cnt2_x2 <- betas[1] + betas[2] * 1 + betas[3] + betas[4]

log(exp(cnt2_x2) / exp(cnt_x2))

betas[3] + betas[4]
```


4. Predict the expected count +/-95% confidence intervals, over the observed range of values x1, assuming x2 is fixed at level 'b'

```{r, echo = T}
data2 <- data.frame(
  x1 = seq(min(dat$x1), max(dat$x1), length.out = 100), 
  x2 = factor(x = rep('b', times = 100)), levels = c('a','b'))

predglm <- predict.glm(object = fit, newdata = data2, tpe = 'link', se.fit = T)
  lower <- exp(predglm$fit - qnorm(0.975) * predglm$se.fit)
  upper <- exp(predglm$fit + qnorm(0.975) * predglm$se.fit)
  
plot(y = exp(predglm$fit), x = data2$x1, xlab = 'x1', ylab = 'expec count', 
     cex.axis = 1.5, cex.lab = 1.5, ylim = c(min(lower), max(upper)), type = 'l')
lines(x = data2$x1, y = lower, lty = 2)
lines(x = data2$x1, y = upper, lty = 2)
```


5. Predict the expected count, ± 95% confidence intervals, of levels "a" and "b", assuming x1 is fixed at
it's mean.

```{r, echo = T}
lvla <- data.frame(x1 = mean(dat$x1), x2 = factor('a'), levels = c('a'))
                  
lvlb <- data.frame(x1 = mean(dat$x1), x2 = factor('b'), levels = c('b'))
       
pa <- predict.glm(object = fit, newdata = lvla, type = 'link', se.fit = T)
respa <- exp(pa$fit)
  lower_a <- exp(pa$fit - qnorm(0.975) * pa$se.fit)
  upper_a <- exp(pa$fit + qnorm(0.975) * pa$se.fit)

pb <- predict.glm(object = fit, newdata = lvlb, type = 'link', se.fit = T)
respb <- exp(pb$fit)
  lower_b <- exp(pb$fit - qnorm(0.975) * pb$se.fit)
  upper_b <- exp(pb$fit + qnorm(0.975) * pb$se.fit)
  
lower_a
upper_a
lower_b
upper_b
  
ab <- data.frame(treatment = factor(c('a','b')), 
                 response = c(respa, respb), 
                 lower = c(lower_a, lower_b), 
                 upper = c(upper_a, upper_b))

library(ggplot2)
graph <- ggplot(ab, aes(treatment, response))
graph + geom_pointrange(aes(ymin = lower, ymax = upper))
```
