---
title: "Homework 6"
author: "Ian Taylor Booth"
date: "April 12, 2021"
output: html_document
---

1. Load data and place into an unmarkedFramePCount object
```{r, echo = T}
library(unmarked)

count <-read.csv("count.csv")
obs_covs <-read.csv("obs_covs.csv")
site_covs <-read.csv("site_covs.csv")


count_mat <-as.matrix(count)
det_covs <-list(js = obs_covs)
nmix_data <-unmarkedFramePCount(y = count_mat,
                                obsCovs = det_covs,
                                siteCovs = site_covs)
```
2. Fit an N-mixture model that assumes conditional detection probability is a function of the detection covariate provided, and expected abundance is a additive function of variables x1 and x2.

```{r, echo = T}

fit <- pcount(formula = ~js ~x1 + x2, 
              data = nmix_data, K = 100)

summary(fit)
```

3. Interpret the e???ect of x1 on the expected count at each site. Verity your interpretation in R

```{r, echo = T}
betas <-coef(fit)
betas

x1 <-betas[1] + betas[2] * 1 + betas[6] + betas[7]
x1ii <-betas[1] + betas[2]*2 + betas[6] +betas[7]

log(exp(x1ii)/exp(x1))

betas[2]

##For every 1 unit increase in x1 there is a .3696 increase in log odds 

```
4. Predict and plot the effect of the supplied detection covariate. Do this over the range of this covariate.
```{r, echo = T}
njs <-data.frame(
  js = seq(from = min(obs_covs$j1, obs_covs$j2, obs_covs$j3),
             to = max(obs_covs$j1, obs_covs$j2, obs_covs$j3), length.out = 100))
  
  
js_min <-min(obs_covs$j1, obs_covs$j2, obs_covs$j3)
js_max <-max(obs_covs$j1, obs_covs$j2, obs_covs$j3)

range = seq(from = js_min, to = js_max, length.out = 100)
  
pred <- predict(object = fit, newdata = njs, type = 'det')
pred 


plot(y = pred$Predicted, 
     x = seq(from = js_min, to = js_max, length.out = 100),
     ylim = c(0,1),
     xlab = 'Detection covariate',
     ylab = 'Detection Probability',
     type = 'l')
lines(x = range, y = pred$lower, lty = 2)
lines(x = range, y = pred$upper, lty = 2)


```
5. Use contrasts to compare expected abundance between all pairwise levels of variable x2. Obtain p-values associated with each contrast and tell me whether you reject or fail to reject each null hypothesis tested.
```{r, echo = T}
pbeta <- betas[1:5]

x <-matrix(
  c(0,0,1,0,0,
    0,0,0,1,0,
    0,0,0,0,1,
    0,0,-1,1,0,
    0,0,-1,0,1,
    0,0,0,-1,1),
  nrow = 6, byrow = T
)

x

con <- x %*% pbeta

vcovcon <- x %*% vcov(fit)[1:5, 1:5] %*% t(x)
vcovcon


secon <- sqrt(diag(vcovcon))
secon

w <-con[,1]/secon
w
pvalues <- 2*pnorm(-1*abs(w))

pvalues

# Null Hypothesis = No change in pairwwise levels of variable x2 for the following levels:
#a/b: fail to reject, a/c: fail to reject, a/d: reject, b/c: fail to reject, b/d: reject, c-d: reject

```
