---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Homework 1 Beach ID's / Years / and Open Beach Area
## fit a fully interactive model that inludes beachID and year


1. Import Beach Data set and examine the 1st few rows
```{r}
BD <- read.csv("beach.csv")
data <- as.data.frame(BD)


#inspect 1st 5 entries
head(BD)
```

2. fit a model that includes interactions between beach ID and year

```{r}
fit_i <- lm(formula = OpenBeach ~ Year * BeachID, data = data)


betas <- coef(fit_i)
betas

betas[1]
betas[2]
betas[3]
betas[4]
betas[5]
betas[6]
```

3. Predict the beach area at all three beaches in 1992

```{r}
#Beach A  
BA = betas[1] + betas[2] * 1992
BA

BA = 362.815

#Beach B
BB = betas[1] + betas[2] * 1992 + betas[3]*1 + betas[5] * 1992 * 1
BB

BB = 100.034


#Beach C
BC = betas[1] + betas[2] * 1992 + betas[4]*1 + betas[6] * 1992 * 1
BC

BC = 214.374

### Answer: In 1992 Beach A was predicted to be 362.815ha , Beach B 100.034ha , and Beach C 214.374
```

4. Plot Predcited Area of each beach aross all years

```{r}
x <- c(1992:2014)

BA2 = betas[1] + betas[2] * x 
BB2 = betas[1] + betas[2] * x + betas[3]*1 + betas[5] * x * 1
BC2 = betas[1] + betas[2] * x + betas[4]*1 + betas[6] * x * 1

BA2
BB2
BC2

BchA <- plot(x = x, y = BA2, type = 'l', xlab = 'year', ylab = 'Pred. Area A')
BchB <- plot(x = x, y = BB2, type = 'l', xlab = 'year', ylab = 'Pred. Area B')
BchC <- plot(x = x, y = BC2, type = 'l', xlab = 'year', ylab = 'Pred. Area C')
```

5. Interpret the rate of change in beach area for each beach

```{r}
#Beach A  
ROC = betas[1] + betas[2] * 1992
ROC1 = betas[1] + betas[2] * 1993

ROC1-ROC
# -18.36491

#Beach B
ROC2 = betas[1] + betas[2] * 1992 + betas[3]*1 + betas[5] * 1992 * 1
ROC3 = betas[1] + betas[2] * 1993 + betas[3]*1 + betas[5] * 1993 * 1

ROC3-ROC2
# -1.322145

#Beach C
ROC4 = betas[1] + betas[2] * 1992 + betas[4]*1 + betas[6] * 1992 * 1
ROC5 = betas[1] + betas[2] * 1993 + betas[4]*1 + betas[6] * 1993 * 1

ROC5-ROC4
# -3.993592

#####Answer
#Beach A is decreasing at a rate of 18.36ha per year
#Beach B is decreasing at a rate of 1.32ha per year
#Beach C is decreasing at a rate of 3.99ha per year
```