---
title: "Homework 8"
author: "Ian Taylor Booth"
date: "4/28/2021"
output: html_document
---

1. Fit an N-mixture model that assumes abundance is a function of wetland size and type, and detection
probability is a function of sky and time

```{r, echo = T}

library(unmarked)
y <- read.csv("sosp_nmix.csv")
site_covs <- read.csv("n_covs_nmix.csv")
p_covs <- read.csv("p_covs_nmix.csv")

det_covs <- list(
  time = data.frame(p_covs[, c('time.1', 'time.2')]),
  sky = data.frame(sky.1 = factor(p_covs$sky.1),
                   sky.2 = factor(p_covs$sky.2))
)

site_covs$herb <- factor(site_covs$herb)
site_covs$shrub1 <- factor(site_covs$shrub1)
site_covs$bareground <- factor(site_covs$bareground)
site_covs$shrub5 <- factor(site_covs$shrub5)

nmix_data <- unmarkedFramePCount (y = as.matrix(y),
                                 siteCovs = site_covs,
                                 obsCovs = det_covs)

fit <- pcount(~ sky + time ~ size + type, nmix_data, K = 100)

```

2. Write a function that calculates the sum of squared Pearson residuals from a fitted model.
```{r, echo = T}

fitstats <- function(fit)
  { 
  obs <- getY(fit) 
  ex <- fitted(fit) 
  det <- predict(fit, type = "det")
  dm <- matrix(det$Predicted, byrow = TRUE, ncol = 2)
  ts <- (obs - (ex)) ^ 2 /  ((ex) * (1 - dm))
  return(sum(ts))
}

```

3. Use the parboot() function in R to simulate the distribution of this test statistic under the assumption
that your fitted model is the data-generating model. Simulate 1000 values of the test statistic. Note
that this may take several minutes

```{r, echo = T}
pb <- parboot(fit, fitstats, nsim=1000)
pb

```

4. Plot the distribution of the simulated test statistic. Include in this plot the value of your test statistic
calculated from your fitted model. What is the null hypothesis you are testing when conducting model
checking? Do you reject or fail to reject this null hypothesis? What are the implications for how well
you model fits the data

Answer: The null hypothesis for model checking is testing whether our model is an accurate fit of the data; or the null hypothesis is checking whether our test statistic is less than our vector of bootstrap samples. H0 : t_B ≥ t0. Because our test statistic is greater than our t_B we reject the null hypothesis. Because we reject the null (our model fits the data) we can say that our fitted model is innacurate and that abudance is not a function of wetland size and type and detection probablity is a not a function of sky and time.

```{r, echo = T}
plot(pb)

```

