---
title: "Homework 8"
author: "Taylor Booth"
date: "April 10, 2021"
output: html_document
---
1. Load data and place into an unmarkedFrameOccu object
```{r, echo = T}

library('unmarked')

bobcat <-read.csv("Bobcat.csv")
head(bobcat)

p_covs <-read.csv("p covariates.csv")
head(p_covs)

psi_covs <-read.csv("psi covariates.csv")
head(psi_covs)

p_covs <-list(
  people = data.frame(p_covs))

occu_data <-unmarkedFrameOccu(y = bobcat,
                              siteCovs = psi_covs,
                              obsCovs = p_covs)
```

2. Fit the following candidate set of models:
detection model  occupancy model
intercept-only   intercept-only
people           intercept-only
intercept-only   disturbance
people           disturbance

 
```{r, echo = T}

library('AICcmodavg')
fit_1 <- occu(~1 ~1, data = occu_data)
fit_2 <- occu(~people ~1, data = occu_data)
fit_3 <- occu(~1 ~Dist_5km, data = occu_data)
fit_4 <- occu(~people ~Dist_5km, data = occu_data)

```

3. Perform model selection with AIC. What is your top model? How do you know? Is there model selection uncertainty?
``` {r, echo = T}

cand.set <- list(
  F1 = fit_1, F2 = fit_2, F3 = fit_3, F4 = fit_4
)

mods <-aictab(cand.set = cand.set, second.ord = F)
head(mods)

```


My top model is model 3; detection set to intercept-only and occupancy set to disturbance. It has the lowest AIC value (Akaike's information criterion), a delta AIC of 0, and a cumulative model weight of 73%. Our selection uncertainty is very low due to the small number of parameters used.



4. Average both the effect of people on detection, and disturbance on occupancy, over all models. Report model-averaged slope coefficients and 95% confidence intervals.

```{r, echo = T}

avg_people <-modavgShrink(cand.set = cand.set,
                          parm = 'people',
                          second.ord = F,
                          parm.type = 'detect')

#mode average and 95% interval
avg_people$Mod.avg.beta
avg_people$Lower.CL
avg_people$Upper.CL

avg_disturbance <-modavgShrink(cand.set = cand.set,
                               parm = 'Dist_5km',
                               second.ord = F,
                               parm.type = 'psi')

##model average slope coefficient and 95% CI for disturbance occupancy 
avg_disturbance$Mod.avg.beta
avg_disturbance$Lower.CL
avg_disturbance$Upper.CL


```

5. Obtain and plot model-averaged predictions of occupancy probability and detection probability. Average over all models, and make predictions over the observed range of each variable.
```{r, echo = T}
##detection probability
summary(occu_data)
min_people <- 0.0
max_people <-5.6

new_df <- data.frame(
  people = seq(from = min_people, to = max_people, length.out = 100)
)

avg_predictions <- modavgPred(cand.set = cand.set,
                              newdata = new_df,
                              second.ord = F,
                              parm.type = 'detect')

avg_predictions

plot(y = avg_predictions$mod.avg.pred,
     x = seq(from = min_people, to = max_people, length.out = 100),
     xlab = "Number of People / 100", ylab = "Detection Probability",
     cex.axis = 1.5, cex.lab = 1.5, type = 'l', ylim = c(0,1))
lines(y = avg_predictions$lower.CL,
      x = seq(from = min_people, to = max_people, length.out = 100),
      lty = 2)
lines(y = avg_predictions$upper.CL,
      x = seq(from = min_people, to = max_people, length.out = 100),
      lty = 2)

##occupancy Probability

new_df2 <- data.frame(
  Dist_5km = seq(from = min(psi_covs$Dist_5km), to = max(psi_covs$Dist_5km), length.out = 100)
)

avg_pred_occu <- modavgPred(cand.set = cand.set,
                              newdata = new_df2,
                              second.ord = F,
                              parm.type = 'psi')

avg_pred_occu

plot(y = avg_pred_occu$mod.avg.pred,
     x = seq(from = min_people, to = max_people, length.out = 100),
     xlab = "Proportion of Area within 5km of Disturbance", ylab = "Occupancy Probability",
     cex.axis = 1.5, cex.lab = 1.5, type = 'l', ylim = c(0,1))
lines(y = avg_pred_occu$lower.CL,
      x = seq(from = min_people, to = max_people, length.out = 100),
      lty = 2)
lines(y = avg_pred_occu$upper.CL,
      x = seq(from = min_people, to = max_people, length.out = 100),
      lty = 2)

```
