---
title: "Homework 2"
author: "Taylor Booth"
date: "February 17th, 2021"
output: html_document
---

```{r, echo = T}
  
beachdata <- read.csv("beach.csv")

head(beachdata)

summary(beachdata)

```
1. This is my answer to question 1:
  
```{r, echo = T}

fit <- lm(OpenBeach ~ Year * BeachID, data = beachdata)
y_obs <- beachdata$OpenBeach
  BA <- unique(beachdata$Year[beachdata$BeachID == 'A'])
  BB <- unique(beachdata$Year[beachdata$BeachID == 'B'])
  BC <- unique(beachdata$Year[beachdata$BeachID == 'B'])
  
betas <- coef(fit)
  betas
  

BAfit = betas[1] + betas[2] * BA
BAfit

BBfit = betas[1] + betas[2] * BB + betas[3]*1 + betas[5] * BB*1
BBfit

BCfit = betas[1] + betas[2] * BC + betas[4]*1 + betas[6] * BC*1
BCfit

fit2 <- c(BAfit, BBfit, BCfit)
fit2

residuals <- y_obs - fit2
residuals

#plot histogram
hist(residuals)


summary(residuals)
summary(fit)


```
Our residuals appear to be reasonably gaussian with some deviation from true gaussian at the 
upped end of the histogram but exhbit a normal distribution.




2. This is my answer to question 2
```{r, echo + T}
summary(fit)

#Coefficient estimates

b0 <- coef(fit)[1]
b0
b1 <- coef(fit)[2]
b1
b2 <- coef(fit)[3]
b2
b3 <- coef(fit)[4]
b3
b4 <- coef(fit)[5]
b4
b5 <- coef(fit)[6]
b5

#estimate standard error
b0SE <- summary(fit)[['coefficients']]['(Intercept)', 'Std. Error']
b0SE

b1SE <- summary(fit)[['coefficients']]['Year', 'Std. Error']
b1SE

b2SE <- summary(fit)[['coefficients']]['BeachIDB', 'Std. Error']
b2SE

b3SE <- summary(fit)[['coefficients']]['BeachIDC', 'Std. Error']
b3SE

b4SE <- summary(fit)[['coefficients']]['Year:BeachIDB', 'Std. Error']
b4SE

b5SE <- summary(fit)[['coefficients']]['Year:BeachIDC', 'Std. Error']
b5SE

#test statistic
b <- 0
test0 <- ((b0 - b)/b0SE)
test0

test1 <- ((b1 - b)/b1SE)
test1

test2 <- ((b2 - b)/b2SE)
test2

test3 <- ((b3 - b)/b3SE)
test3

test4 <- ((b4 - b)/b4SE)
test4

test5 <- ((b5 - b)/b5SE)
test5


#verify by comparison
 summary(fit)

test0
test1
test2
test3
test4
test5


```
3. This is my answer to question 3:
```{r, echo = T}

DF <-  nrow(beachdata) - 6

#calc pvalues

pv0 <- pt(-1 * abs(test0), df = DF) + (1 - pt(abs(test0), df = DF))
pv0

pv1 <- pt(-1 * abs(test1), df = DF) + (1 - pt(abs(test1), df = DF))
pv1
pv2 <- pt(-1 * abs(test2), df = DF) + (1 - pt(abs(test2), df = DF))
pv2
pv3 <- pt(-1 * abs(test3), df = DF) + (1 - pt(abs(test3), df = DF))
pv3
pv4 <- pt(-1 * abs(test4), df = DF) + (1 - pt(abs(test4), df = DF)) 
pv4
pv5 <- pt(-1 * abs(test5), df = DF) + (1 - pt(abs(test5), df = DF)) 
pv5


#verify
summary(fit)
pv0
pv1
pv2
pv3
pv4
pv5

```
   The null hypothesis is that our betas will equal zero

H0: b0=0, b1=0, b2=0, b3=0, b4=0, b5=0

Ha: b0 /=0, b1 /=0, b2/=0, b3/=0, b4/=0, b5/=0

We reject the null hypothesis all of the p-values fall below our alpha value (0.05). P-value = 0.000; all intercepts


4. This is my answer to question 4:
  
Null hypothesis 

h0: b_rn = -60
ha: b_rn > -60
  
``` {r, echo + T}

  # Test Statistic                   

testi <- (coef(fit)[2] - 60) / summary(fit) [['coefficients']]['Year', 'Std. Error']
testi

# P Value
DF
pt(-1 * abs(testi), df = 56)


```
With an alpha value = 0.05, we reject the null hypothesis (p-value = 0.0000, t-stat = -61.62812, df = 56)



5. This is my answer to question 5:
  
``` {r, echo + T}
  
#Beach A  
ROC = betas[1] + betas[2] * 1992
ROC1 = betas[1] + betas[2] * 1993

ROC1-ROC
= -18.36491

#Beach B
ROC2 = betas[1] + betas[2] * 1992 + betas[3]*1 + betas[5] * 1992 * 1
ROC3 = betas[1] + betas[2] * 1993 + betas[3]*1 + betas[5] * 1993 * 1

ROC3-ROC2
= -1.322145

#Beach C
ROC4 = betas[1] + betas[2] * 1992 + betas[4]*1 + betas[6] * 1992 * 1
ROC5 = betas[1] + betas[2] * 1993 + betas[4]*1 + betas[6] * 1993 * 1

ROC5-ROC4
= -3.993592

```
#####Answer
#Beach A is decreasing at a rate of 18.36ha per year
#Beach B is decreasing at a rate of 1.32ha per year
#Beach C is decreasing at a rate of 3.99ha per year
  
