---
title: "Homework 3"
author: "Taylor Booth"
date: "March 3, 2021"
output: html_document
---

```{r, echo = T}
dat <- read.csv("Homework 3 Data.csv")
head(dat)
```

1. This is my answer to question 1
```{r, echo = T}
fit <- glm(y ~ x1 + x2, family = binomial, data = dat)
summary(fit)
```

2. This is my answer to question 2
```{r, echo = T}
betas <- coef(fit)
x1 <- plogis(betas[1] + betas[2]*0 + betas[3]*0)
x1i <- plogis(betas[1] + betas[2]*1 + betas[3]*0)

log((x1i / (1 -x1i)) / (x1 / (1-x1)))
```
# -0.167976
for every 1 unit increase of x1 there is a -0.167976 change in the log odds of success


3. This is my answer to question 3
```{r, echo = T}
x2 <- plogis(betas[1] + betas[2]*0 + betas[3]*0)
x2i <- plogis(betas[1] + betas[2]*0 + betas[3]*1)

log((x2i / (1-x2i)) / (x2 / (1-x2)))
```
#-.9679244

for every 1 unit increase of x2 there is a -0.9679244 change in the log odds of success


4. This is my answer to question 4
```{r, echo = T}
#### Wald Test x1
tx1 <- betas[2] / summary(fit)[['coefficients']]['x1', 'Std. Error']
summary(fit)[['coefficients']]['x1', 'z value']

#### p-value
2 * pnorm(-1 * abs(tx1), mean = 0, sd = 1)
summary(fit)[['coefficients']]['x1', 'Pr(>|z|)']

####wald Test x2b
tx2b <- betas[3] / summary(fit)[['coefficients']]['x2b', 'Std. Error']
summary(fit)[['coefficients']]['x2b', 'z value']

# p-value
2 * pnorm(-1 * abs(tx2b), mean = 0, sd = 1) 
summary(fit)[['coefficients']]['x2b', 'Pr(>|z|)']
```
The p value for x1 is 0.6451594 at a significance level of 0.05 we fail to reject the null hypothesis

The p value for x2 is 0.01926485 at a significance level of 0.05 we reject the null hypothesis



5. This is my answer to question 5
```{r, echo = T}
xv <- seq(from = min(dat$x1), to = max(dat$x1), length.out = 100)
y <- betas[1] + betas[2] * xv
plot(x=xv, y = plogis(y), ylab = 'Prob Success', xlab = 'x1', 
     cex.axis = 1.5, cex.lab = 1.5, type = 'l')
```
