---
title: "Homework 5"
author: "Taylor Booth"
date: "April 4, 2021"
output: html_document
---

1. Load data and place into an unmarkedFrameOccu object

```{r, echo = T}
bobcat <- read.csv("Bobcat.csv")

pcovar <- read.csv("p covariates.csv")
psicovar <- read.csv("psi covariates.csv")

library(lattice)
library(unmarked)

peoplecov <- list(pcovar = data.frame(pcovar))
head(peoplecov)
class(peoplecov[1])

occudat <- unmarkedFrameOccu(y = bobcat, siteCovs = psicovar, obsCov = peoplecov)
```


2. Fit an occupancy model that assumes conditional detection probability is a function of the number of
people photographed at a site, and occupancy probability is a function of disturbance within 5km.

```{r, echo = T}
fit <- occu(formula = ~pcovar ~ Dist_5km, data = occudat)
summary(fit)
class(psicovar)

```

3. Interpret the effect of the number of people photographed at each site on conditional detection probability
of bobat. Verity your interpretation in R.

Answer: The number of people photgraphed has a change in log odds of -0.058 on the conditional detection probablity of bobcat

```{r, echo = T}
betas <- coef(fit)
betas

people1 <- betas[3] + betas[4] * 1
people2 <- betas[3] + betas[4] * 2

log(plogis(people2)/ plogis(people1))

summary(fit)
```

4. Re-create the Wald test for the effect of disturbance within 5km. What is the null hypothesis? Do you
reject or fail to reject this null hypothesis?

Answer: The null hypothesis is that there is no change in log odds of detection when disturbance is within 5km. We reject the null hypothesis since our p value is
so small (p = 1.318e-15)

```{r, echo = T}
wald <- betas[3] / .377
wald

2 * pnorm(-1 * abs(wald))


```

5. Predict and plot the effect of disturbance within 5km on bobcat occupancy probability. Do this over
the range of observed disturbance within 5km.

```{r, echo = T}
max(psicovar$Dist_5km)
newdat <- data.frame(Dist_5km = seq(from=min(psicovar$Dist_5km), to = max(psicovar$Dist_5km), length.out = 100), 
                  people = rep(0, lenth.out=100))
newdat

pred <- predict(object = fit, newdata = newdat, type = "state", appendData = TRUE)
pred
```
